---
title: "Eurostat database and package"
output: html_document
---

### `{eurostat}` 

`{eurostat}` R package provides tools to access open data from [Eurostat](http://ec.europa.eu/eurostat): 

* data search, 
* download, 
* manipulation, 
* visualization. 


### Installation


```{r install, eval=FALSE, echo=T}
# Note the "eval=FALSE" argument: {r} chunk is not run.
install.packages("eurostat")
```


--- 

## Using the package

* [Cheat sheet: eurostat R package](http://ropengov.github.io/eurostat/articles/cheatsheet.html)
* [Tutorial (vignette) for the eurostat R package](http://ropengov.github.io/eurostat/articles/eurostat_tutorial.html)
* [Detailed function documentation](http://ropengov.github.io/eurostat/reference/index.html)

--- 

| Command              | Description                                          |
|----------------------|------------------------------------------------------|
| `get_eurostat_toc()` | Download table of contents of Eurostat datasets     |  
| `search_eurostat()`  | Retrieve (grep) datasets titles from Eurostat                   |  
| `get_eurostat()`     | Read Eurostat data                                   |  
|                      |                                                      |
| `label_eurostat()`   | Get Eurostat code descriptions                       |  
|                      |                                                      |
| `get_eurostat_geospatial()`   | Download geospatial data from GISGO         |  


--- 

## Search Eurostat for data

```{r, echo=T, message=F}
options(readr.default_locale=readr::locale(tz="Europe/Berlin"))
require(eurostat)
require(knitr)
require(tidyr)
require(dplyr)
require(ggplot2)
require(RColorBrewer)
```


```{r, eval=FALSE}
# To actually run this {r} chunk, change the eval argument
toc <- get_eurostat_toc() # Downloads Table of Contents of Eurostat Data Sets
class(toc)
dim(toc)
str(toc,list.len = 10) # only few items listed
```

With `search_eurostat()`, you can search the table of contents for particular patterns. 

* *regex*: R regular expression syntax is used

* e.g. we search for Unemployment at NUTS 2 level.


```{r}
# kable() generates tabular (formatted) output in Rmd files
kable(search_eurostat(".*unemployment.*rates.*NUTS", fixed=F))
kable(search_eurostat(".*Unemployment.*rates.*NUTS", fixed=F))
```

```{r, eval=FALSE, echo=T}
## Note that search_eurostat() is case-sensitive.
##
## Alternatively, you can use grep() to search a downloaded TOC
toc <- get_eurostat_toc() # Downloads Table of Contents of Eurostat Data Sets
toc[grep(".*unemployment.*rates.*NUTS", ignore.case = T, toc$title),]

```


--- 


## Download data

As na example, let's choose the **Unemployment rates by sex, age and NUTS 2 regions (%)** dataset	`lfst_r_lfu3rt`

* All datasets are available through a web browser (see the last string in the web address)
* http://appsso.eurostat.ec.europa.eu/nui/show.do?dataset=lfst_r_lfu3rt

Download the data:

```{r, message=F}
Un.DF <- get_eurostat("lfst_r_lfu3rt", time_format = "num") # note the simplified time format
dim(Un.DF)
# (5 age groups) * (3 gender categories) * (19 year 1999-2017) * (504 geo units) = 143.640
str(Un.DF)
summary(Un.DF) # note that observations are annual
head(Un.DF,10)
```


By default, variable identification is provided through Eurostat codes.

* To get human-readable labels instead, we can use `label_eurostat()` function
* Good for orientation in the dataset, NOT for `gather()` , `spread()` data handling


```{r, echo=T, message=F}
Un.DF.l <- label_eurostat(Un.DF, fix_duplicated = T)
```

```{r}
head(Un.DF.l,10)
summary(Un.DF.l)
```

Also, codes and their "descriptions" can be shown side-by-side:

```{r}
head(kable(cbind(as.character(unique(Un.DF$geo)),as.character(unique(Un.DF.l$geo)))),17)
```

---

## Data handling

We can simply save the data for subsequent use:


```{r}
write.csv(Un.DF, "datasets/Unemployment.csv", row.names = F)
```

----

We can use `{tidyverse}` and `{ggplot2}` functions to filter and plot data. 

**Example 1: **

* `Y15-74` i.e. age group *from 15 to 74 years*  
* Select data 2010 and newer  
* Total unemployment only (no M/F/T) structure  
* Select only: Austria, Czech Republic, Germany, Hungary, Poland, Slovakia  
* NUTS 0 (State-level)  



```{r}
Un.DF %>% 
  filter(age == "Y15-74", time >= 2010, sex == "T") %>%
  subset(geo %in% c("AT","CZ","DE","HU","PL","SK")) %>% 
  ggplot(aes(x = time, y = values, colour = geo))+
    geom_line()+
    ggtitle("Unemployment rates")+
    ylab("Unemployment (%)")+
    theme_minimal()
```

--- 

**Example 2: **

* `Y15-74` i.e. age group *from 15 to 74 years*  
* Select data for 2017  
* Total unemployment only (no M/F/T)  
* NUTS 2  
* Austria, Czech Republic, Germany, Hungary, Poland, Slovakia  
* Draw choropleth (cartogram, infomap)

```{r, echo=T, message=F}
# Download geospatial data from GISCO # NUTS revisions: e.g.: 2010, 2013, 2016.
geodata <- get_eurostat_geospatial(resolution = "60", nuts_level = "2", year = 2016)
# Filter Unemployment dataset
Un.2016 <- Un.DF %>% 
  filter(age == "Y15-74", time == 2017, sex == "T") %>% # filter variables, year, sex
  filter(nchar(as.character(geo)) == 4) %>% # NUTS2 regions have a 4-digit id
  mutate(NUTS0 = substr(as.character(geo), start=1, stop=2)) %>% # retrieve NUTS0 id from NUTS2
  subset(NUTS0 %in% c("AT","CZ","DE","HU","PL","SK")) 
# Join Unemployment data with "map data"
map_data <- inner_join(geodata, Un.2016)
# plot the data
ggplot()+ 
  geom_sf(data = map_data, aes(fill = values))+
  scale_fill_gradientn('Unem. \n(%) ', colours=brewer.pal(8, "Oranges"))+
  ggtitle("Unemployment rates")
```

---

**Quick assignment:** Add Netherlands (`NL`) to both plots (Examples 1 & 2).

--- 