---
title: "Krigging (and IDW) example - Meuse river sediments"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sp)
library(gstat)
library(dplyr)
library(ggplot2)
library(scales) # for "comma"
library(magrittr)
data(meuse)
data(meuse.grid)
```

We shall be predicting zinc-sediment levels. Exampble is based on the Meuse dataset:

```{r}
glimpse(meuse)
glimpse(meuse.grid) # empty grid - zinc levels will be predicted here
```

We can use `ggplot2` to visually inspect how zinc varies over the domain of interest where we map concentration to point size:

```{r}
meuse %>% as.data.frame %>% 
  ggplot(aes(x, y)) + geom_point(aes(size=zinc), color="blue", alpha=3/4) + 
  ggtitle("Zinc Concentration (ppm)") + coord_equal() + theme_bw()
```


----

Prepare data for spatia analysis (IDW, krigging)


```{r}
coordinates(meuse) = ~x+y
coordinates(meuse.grid) = ~x+y
```

----

## IDW

```{r}
idwmodel <- idw(log(zinc) ~1, meuse,meuse.grid,
               maxdist = Inf, idp = 1) 
idwmodel %>% as.data.frame %>%
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  scale_x_continuous(labels=comma) + scale_y_continuous(labels=comma) +
  theme_bw()
```

* Voronoi-type diagram

```{r}
idwmodel2 <- idw(log(zinc) ~1, meuse,meuse.grid,
               maxdist = Inf, idp = 20) 
idwmodel2 %>% as.data.frame %>%
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  scale_x_continuous(labels=comma) + scale_y_continuous(labels=comma) +
  theme_bw()
```

----

## Krigging


* 1) Calculate the sample variogram. This is done with the `variogram()` function.

* 2) Fit a model to the sample variogram using `fit.variogram()` function

* 3) Use the fitted variogram for krigging ($\lambda_i$) weights, using `krige()` function

* 4) Plot the krigged data using `ggplot()`

```{r}
# 1
lzn.vgm <- variogram(log(zinc)~1, meuse) # calculates sample variogram values 
# 2 
lzn.fit <- fit.variogram(lzn.vgm, model=vgm(psill=1, "Sph", range=900, nugget=1))
# 3
lzn.kriged <- krige(log(zinc) ~ 1, meuse, meuse.grid, model=lzn.fit)
# 4
lzn.kriged %>% as.data.frame %>%
  ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=var1.pred)) + coord_equal() +
  scale_fill_gradient(low = "yellow", high="red") +
  scale_x_continuous(labels=comma) + scale_y_continuous(labels=comma) +
  theme_bw()
```



----

#### Quick exercise:

* Repeat the krigging process (including plotting) while fitting the empirical semivariogram to a  Gaussian curve.


```{r}
# 1

# 2 

# 3

# 4

```



----- 

For application to London's house prices, see [example here](https://rpubs.com/chrisbrunsdon/gwdplyr)
